---
title: "JHU_DS_MilestoneReport"
author: "Ghostdatalearner"
date: "17/03/2015"
output: html_document
---

# Introduction
This document is a report on the preliminary analysis of text source files with the final goal of building a word prediction application. It is part of the capstone project of [John Hopkins University Data Specialization series at Coursera<https://www.coursera.org/specialization/jhudatascience/>. 

# Executive Report

# Environment and Data
Analysis has been performed on a personal computer running Ubuntu 14.04LTS,

The data set is provided by SwiftKey and was downloaded from <https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip>. It contains twelve files, with samples of texts in Finnish, German, English and Russian taken from Twitter, news services and blogs. We have worked with the English files.

## Raw Data Summary
These are the statistics of the three raw files, that we have got with the shell commands ls -l,  wc -l and wc -w

```html
File                Size (bytes)         Lines              Words
=======================================================================
en_US.blogs.txt     210160014             899288           37334114
en_US.news.txt      205811889            1010242           34365936
en_US.twitter.txt   167105338            2360148           30359804

```


```{r, echo=FALSE}
library(ggplot2)
library("R.utils")
options(warn=-1)
```

## Sampling
As files are rather heavy, we decided to work with a representative sample to make the exploratory analysis. We pick at randmon a 20% of the lines of each file.

```{r}
set.seed(1234)
#filename <- c("en_US.twitter.txt","en_US.news.txt","en_US.blogs.txt")
filename <- c("en_US.news.txt")
sampletext <- c()
for (j in filename)
{
  cl <- countLines(j)
  sampleindex <- sample(seq(1,cl),round(0.02*cl))
  print(paste("file:",j," Number of lines:",cl))
  con <- file(description=j, open="r")
  for(i in 1:cl) {
    tmp <- readLines(con, n=1)
    if (length(which(sampleindex == i))>0)
      sampletext <- c(sampletext,tmp)   
  }
  close(con)
}
```


# Expl


